Beyond the best guess: Distributional predictions from large language models for human forced-choice data

### 2023-10-06 03-paper/paper-LLM-probmod.tex
revised §1

The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
% especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
Recently an increasing number of diverse applications has gone beyond using LLMs out-of-the-box, i.e., based on their single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
Examples aimed at improving output quality include sophisticated prompting \citep{LiuLiu2022:Generated-Knowl}, structured reasoning or ``neuro-symnolic AI'' \citep{CreswellShanahan2022:Selection-Infer}, or applications of LLMs as scoring functions in open-ended applications \citep{ZhangLehman2023:OMNI:-Open-ende}.
Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep{WongGrand2023:From-Word-Model}.
\mf{enlarge briefly, include more references}

For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
The prevalent approach towards understanding the performance of LLMs relies on benchmark testing which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``true answer'' exists, averaged over many instances of this task.
Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
There is also some work investigating whether LLMs can make adequate \emph{quantitative predictions}.
For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.

This work seeks to extend the investigation of the human-likeness of LLM predictions even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.

\begin{itemize}
  \item distributional predictions are more informative, so beneficial for ``understanding'' and for ``applications''
  \item for ``understanding LLMs'' more informative, richer predictions can be more diagnostic about LLM capabilities (risks and potentials)
  \item reliable full-distributional predictions from LLMs can be useful in applications, e.g., relative scoring based on elusive intuitive concepts like ``relevance'' or ``interestingness''
\end{itemize}

>>>>

The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
% especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
Examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
Information from LLMs is also used  to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.

For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.

This work seeks to extend the investigation of the human-likeness of quantitative information derived from LLMs even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.

A major contribution of this work is methodological.
By embedding numerical scores from LLMs into probabilistic models for human multiple-choice data, we can use Bayesian data analysis to train, test and compare different LLM-derived predictor scores.
Concretely, after motivating the desire to assess quantitative information from LLMs in Section~\ref{motivation}, Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though a condition-level predictor from the RSA model is.
Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the task- or condition-level and compares them against the human data and each other.
We find that not every way of constructing condition-level predictions from item-level scores is equally good.
\mf{grant conclusion?}
